Hunting the Parent of the Orphan Stream: Identifying Stream Members \\ from Low-Resolution Spectroscopy

### 2012-10-16 2012-oss-mrs/Response_to_referee_report.txt

Detailed comments follow.

Title

1) ``Host' is not the usual way of referring to a stream's progenitor. Since stars in the stream are unbound from the progenitor and separated by kiloparsecs and billions of years, it is better to say ``Origin of the Orphan Stream' or ``Parent of the Orphan Stream' or ``Orphan Stream Progenitor'.

Response:
Altered title to "Hunting the Parent of the Orphan Stream: Identifying Stream Members from Low-Resolution Spectroscopy", and two instances in the text where the Orphan Stream progenitor was referred to as the host:

@@

 - Section 1
   "as well as make predictions about the nature of the host system."

 - Section 5,
   "help determine both the nature of the host system before it's discovered"

In both cases, 'host system' has been changed to 'progenitor'.

Response:
Manuscript edited as requested. Following sentence edited to minimise repeated words. Text now reads as 'Stars which were once gravitationally bound to the satellite are strewn along the progenitor's orbit in leading and trailing streams of stars. The kinematics of the stream..'

@@

4) In the last paragraph, it would better read ``recover stream giants and contaminating foreground dwarfs.'

Response:
Text altered as requested.

5) In this same paragraph, you should specify why no 2mass photometry is available. One assumes this is because even the giants are too faint to be detected in the 2mass survey.

Response:
One assumes correctly. Text altered to read 'Given our colour selection we expect to recover giants and contaminating dwarf contaminants. Although the 2MASS $JHK$ colours can help to separate dwarfs and giants, our target K-giants stars are too faint to be detected in the 2MASS catalogue.'

@@

Section 4.2

9) In the first paragraph, it is not clear from Figure 2 why this
particular separation line is appropriate or optimal. More explanation and references should be provided. Also, ``allowing multiple inclusive cuts' is somewhat confusing. Should this be ``imposing simultaneous cuts on several observables'?

10) In the last paragraph, the selection criteria used by Casey et al. (2012a) should be briefly summarized here.

Response:
Text now reads as 'This fraction of giants ($\sim$44\%) is much larger than that found by \citeauthor{Casey_et-al_2012} \citeyear{Casey_et-al_2012}a who employed a similar colour selection.'

@@

13) Given the small number and significant color and magnitude spread among the highly probable members about the best-fitting isochrone in Figure 4, a distance uncertainty of only 1 kpc seems unlikely. How was this uncertainty estimated?

Response:

14) The last sentence should include ``21.4 +/- 1 kpc estimated by Grillmair (2006) and Newberg et al. (2010).', since the distance estimates in these papers were independent and complementary, one using main sequence fitting and the other using BHB stars.

Response:
Text altered to 'Our derived distance is in reasonably good agreement with the $21.4 \pm 1.0$\,kpc independently deduced by \citet{Grillmair_2006} and \citet{Newberg_et-al_2010}.'

### 2012-10-16 2012-oss-mrs/ms.tex
Half of the issues resolved

Figure \ref{fig:velocities} shows a histogram of our galactocentric velocities, compared to the predicted smooth line-of-sight velocity distribution for this region from the Besan\c{c}on model \citep{Robin_et-al_2003}. We have selected predicted stars from the Besan\c{c}on model using the same criterion as outlined in \S\ref{sec:target-selection} after using the \citet{Jordi_et-al_2006} colour transformations. It is clear that our target selection has yielded mostly nearby thick disk dwarf stars with $V_{GSR} \approx -120$\,km s$^{-1}$. The expected kinematic signature for the Orphan Stream, as \citet{Newberg_et-al_2010} reported, is labelled in Figure \ref{fig:velocities}. There is no obvious sharp kinematic peak representative of the Orphan Stream in our sample. From kinematics alone, our targets appears largely indistinguishable from a smooth halo distribution.

@@

We have measured the equivalent width of the gravity-sensitive Mg\,\textsc{I} line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia_Starkenburg_2012}. At a given temperature (or $g - r$) and metallicity, giant stars present narrower Mg\,\textsc{I} absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In some cases no Mg\,\textsc{I} 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criterion alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS de-reddened\footnote{All magnitudes presented in this letter are de-reddened using the \citet{Schlegel_Finkbeiner_Davis_1998} dust maps.} $g - r$, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criterion is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

@@

Our dwarf/giant separation line in Figure \ref{fig:ew-mg} yields 425 potential giants. This fraction of giants ($\sim$44\%) is much larger than what would be expected given our selection criterion (e.g. see \citeauthor{Casey_et-al_2012} \citeyear{Casey_et-al_2012}a). Hence, our gravity criterion is extremely inclusive and susceptible to dwarf contamination. Upon taking the intersection of our kinematic and gravity selections, we deduce eighteen stars that appear likely Orphan Stream K giants.

@@

We have measured the equivalent widths of the Ca\,\textsc{II} near-infrared triplet lines for stars that meet our kinematic and surface gravity criteria. After correcting for luminosity, the equivalent width of the Ca\,\textsc{II} triplet lines provide a good indication of the overall metallicity of a RGB star \citep{Amandroff_Da_Costa_1991}. We have employed the \citet{Starkenburg_et-al_2010} relationship and corrected for luminosity in $g$ against the horizontal branch magnitude at $g_{HB}$ = 17.1 \citep{Newberg_et-al_2010}. Strictly speaking, the Ca\,\textsc{II}\---[Fe/H] calibration is only valid for stars brighter than the horizontal branch, although the relationship only becomes significantly inappropriate near $g \-- g_{HB} \sim +1$ \citep{Saviane_et-al_2012}. Many of our candidates are fainter than this valid luminosity range, and therefore they should not be excluded solely because of their derived metallicities. Stars fainted than $g_{HB}$ will have slightly lower metallicities than predicted by our Ca\,\textsc{II}\---[Fe/H] relationship, and for these stars we will only use metallicities to assign a relative qualitative likelihood for stream membership.
Given a distance to the Orphan Stream, we can also estimate a star's metallicity through isochrone fitting. We have used a 10\,Gyr \citet{Girardi_et-al_2008} isochrone at 21.4\,kpc \citep{Newberg_et-al_2010} and found metallicities for all eighteen likely stream members from their best-fitting isochrone. Derived metallicities from Ca\,\textsc{II} line strengths and isochrone fitting that are consistent (within $\pm0.3$\,dex) indicates these measurements are reliable, and that these stars are indeed at a distance of $\sim$21.4\,kpc. We find ten highly likely stream members with consistently derived metallicities. They fall within the shaded region illustrated in Figure \ref{fig:feh}.

@@

Given that our Orphan Stream giants cover a wide evolutionary range along the giant branch (Figure \ref{fig:cmd}), we are in a good position to revise the distance estimate to the stream. Given a 10\,Gyr \citet{Girardi_et-al_2008} isochrone at $[\mbox{Fe/H}] = -1.63$, we find a best-fitting distance to the stream of $22.5 \pm 1.0$\,kpc at $(l, b) = (250\,^\circ,\,50\,^\circ)$. This isochrone is shown in Figure \ref{fig:cmd}, illustrating that three of our nine most probable member stars lie perfectly on this isochrone. Our derived distance is in reasonably good agreement with the $21.4 \pm 1.0$\,kpc deduced by \citet{Newberg_et-al_2010}.

@@

We have presented a detailed analysis to isolate individual Orphan Stream K giants from low-resolution spectroscopy using a combination of photometric, kinematic, gravity, metallicity, and proper motion information. Although each individual criterion is likely to induce some level contamination, their intersection reveals nine highly probable, self-consistent, Orphan Stream K giants.  We deduce a median stream metallicity of $[\mbox{Fe/H}] = -1.63 \pm 0.19$ and find an intrinsically wide metallicity spread of $\sigma = 0.56$\,dex, indicative of a dSph origin. Unlike other stellar tracers, K-type giants can exist at all metallicities, hence our derived metallicity spread is likely representative of the true stream metallicity distribution function. Recall that the metallicity determination was performed after kinematic and gravity cuts, and three of our most probable members lay perfectly on a 10\,Gyr isochrone of $[\mbox{Fe/H}] = -1.63$. Our stream members indicate a distance to the stream of $22.5 \pm 1.0$\,kpc at $(l, b) = (250\,^\circ,\,50\,^\circ)$, in agreement with that deduced by \citet{Newberg_et-al_2010}.

>>>>

Figure \ref{fig:velocities} shows a histogram of our galactocentric velocities, compared to the predicted smooth line-of-sight velocity distribution for this region from the Besan\c{c}on model \citep{Robin_et-al_2003}. We have selected predicted stars from the Besan\c{c}on model using the same criteria as outlined in \S\ref{sec:target-selection} after using the \citet{Jordi_et-al_2006} colour transformations. It is clear that our target selection has yielded mostly nearby thick disk dwarf stars with $V_{GSR} \approx -120$\,km s$^{-1}$. The expected kinematic signature for the Orphan Stream, as \citet{Newberg_et-al_2010} reported, is labelled in Figure \ref{fig:velocities}. There is no obvious sharp kinematic peak representative of the Orphan Stream in our sample. From kinematics alone, our targets appears largely indistinguishable from a smooth halo distribution.

@@

We have measured the equivalent width of the gravity-sensitive Mg\,\textsc{I} line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia_Starkenburg_2012}. At a given temperature (or $g - r$) and metallicity, giant stars present narrower Mg\,\textsc{I} absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In some cases no Mg\,\textsc{I} 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criteria alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS de-reddened\footnote{All magnitudes presented in this letter are de-reddened using the \citet{Schlegel_Finkbeiner_Davis_1998} dust maps.} $g - r$, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criteria is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

@@

Our dwarf/giant separation line in Figure \ref{fig:ew-mg} yields 425 potential giants. This fraction of giants ($\sim$44\%) is much larger than that found by \citeauthor{Casey_et-al_2012} \citeyear{Casey_et-al_2012}a who employed a similar colour selection. Hence, our gravity criteria is extremely inclusive and susceptible to dwarf contamination. Upon taking the intersection of our kinematic and gravity selections, we deduce eighteen stars that appear likely Orphan Stream K giants.

@@

We have measured the equivalent widths of the Ca\,\textsc{II} near-infrared triplet lines for stars that meet our kinematic and surface gravity criteria. After correcting for luminosity, the equivalent width of the Ca\,\textsc{II} triplet lines provide a good indication of the overall metallicity of a RGB star \citep{Amandroff_Da_Costa_1991}. We have employed the \citet{Starkenburg_et-al_2010} relationship and corrected for luminosity in $g$ against the horizontal branch magnitude at $g_{HB}$ = 17.1 \citep{Newberg_et-al_2010}. Strictly speaking, the Ca\,\textsc{II}\---[Fe/H] calibration is only valid for stars brighter than the horizontal branch, although the relationship only becomes significantly inappropriate near $g \-- g_{HB} \sim +1$ \citep{Saviane_et-al_2012}. Many of our candidates are fainter than this valid luminosity range, and therefore they should not be excluded solely because of their derived metallicities. Stars fainter than $g_{HB}$ will have slightly lower metallicities than predicted by our Ca\,\textsc{II}\---[Fe/H] relationship, and for these stars we will only use metallicities to assign a relative qualitative likelihood for stream membership.
Given a distance to the Orphan Stream, we can also estimate a star's metallicity through isochrone fitting. We have used a 10\,Gyr \citet{Girardi_et-al_2008} isochrone at 21.4\,kpc \citep{Newberg_et-al_2010} and found metallicities for all eighteen likely stream members from their best-fitting isochrone. Derived metallicities from Ca\,\textsc{II} line strengths and isochrone fitting that are consistent (within $\pm0.3$\,dex) indicates these measurements are reliable, and that these stars are indeed at a distance of $\sim$21.4\,kpc. We find ten highly likely stream members with consistently derived metallicities. They fall within the shaded region illustrated in Figure \ref{fig:feh}.

@@

Given that our Orphan Stream giants cover a wide evolutionary range along the giant branch (Figure \ref{fig:cmd}), we are in a good position to revise the distance estimate to the stream. Given a 10\,Gyr \citet{Girardi_et-al_2008} isochrone at $[\mbox{Fe/H}] = -1.63$, we find a best-fitting distance to the stream of $22.5 \pm 1.0$\,kpc at $(l, b) = (250\,^\circ,\,50\,^\circ)$. This isochrone is shown in Figure \ref{fig:cmd}, illustrating that three of our nine most probable member stars lie perfectly on this isochrone. Our derived distance is in reasonably good agreement with the $21.4 \pm 1.0$\,kpc independently estimated by \citet{Grillmair_2006} and \citet{Newberg_et-al_2010}.

@@

We have presented a detailed analysis to isolate individual Orphan Stream K giants from low-resolution spectroscopy using a combination of photometric, kinematic, gravity, metallicity, and proper motion information. Although each individual criteria is likely to induce some level contamination, their intersection reveals nine highly probable, self-consistent, Orphan Stream K giants.  We deduce a median stream metallicity of $[\mbox{Fe/H}] = -1.63 \pm 0.19$ and find an intrinsically wide metallicity spread of $\sigma = 0.56$\,dex, indicative of a dSph origin. Unlike other stellar tracers, K-type giants can exist at all metallicities, hence our derived metallicity spread is likely representative of the true stream metallicity distribution function. Recall that the metallicity determination was performed after kinematic and gravity cuts, and three of our most probable members lay perfectly on a 10\,Gyr isochrone of $[\mbox{Fe/H}] = -1.63$. Our stream members indicate a distance to the stream of $22.5 \pm 1.0$\,kpc at $(l, b) = (250\,^\circ,\,50\,^\circ)$, in agreement with that deduced by \citet{Grillmair;et-al_2006} and \citet{Newberg_et-al_2010}.
